Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-10-11 SEIR.tex
EC fitting details

Some behavioural changes (\eg mask wearing) are not captured by the contact survey.
To allow for these changes, $\beta$ is allowed to vary over time.
In particular, $\beta$ follows a random walk in log space on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{w(t)} \\
    \beta_1 &= 1 \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim N(0, 1)
\end{align}
where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.

@@

\Textcite{osogamiClosed} prove that EC distributions have the following properties:
* Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic
phase-type distribution can be well-matched for some $P$
which is a EC distribution.
* The number of phases used by $P$ is at most one more than the optimal number.
* The parameters of $P$ are available in closed form, and the expression for these is derived.

@@

@@

\begin{figure}
\begin{tikzpicture}[
    node distance = 2cm
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

    \node[state] (S) {$s$};
    \node[state, right of=S] (E1) {$e_1$};
    \node[state, right of=E1] (E2) {$e_2$};
    \node[state, right of=E2] (I1) {$i_1$};
    \node[state, right of=I1] (I2) {$i_2$};
    \node[state, right of=I2] (R) {$r$};

    \path (S) edge node {$\beta$} (E1)
          (E1) edge node {$\sigma$} (E2)
          (E2) edge node {$\sigma$} (I1)
          (I1) edge node {$\gamma$} (I2)
          (I2) edge node {$\gamma$} (R);

    \node[state, below=3cm of E1] (P0) {P0};
    \node[state, right of=P0] (P1) {P1};
    \node[state, right of=P1] (P2) {P2};
    \node[state, right of=P2] (P3) {P3};
    \node[state, right of=P3, draw=none] (Pdots) {...};

    \path (P0) edge (P1)
          (P1) edge (P2)
          (P2) edge (P3);

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
  \caption{SEIR model}
\end{figure}

>>>>

Some behavioural changes (\eg mask wearing) are not captured by the contact survey.
To allow for these changes, $\beta$ is allowed to vary over time.
In particular, $\beta$ follows a random walk in log space on the weekly timescale:
\begin{align}
    \beta(t) &= \beta_{w(t)} \\
    \beta_1 &= 1 \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $w(t)$ is the week number of time $t$, with week 1 being the first week modelled.

@@

\Textcite{osogamiClosed} prove that EC distributions have the following properties:
\begin{itemize}
    \item Any distribution that can be well-matched (defined as having the same first three moments) by an acyclic phase-type distribution can be well-matched for some $P$ which is an EC distribution.
    \item The number of phases used by $P$ is at most one more than the optimal number.
    \item The parameters of $P$ are available in closed form, and the expression for these is derived.
\end{itemize}

@@

I use the mapfit package~\autocite{mapfit} to compute $P$ for the posterior mean of the distribution estimated in \cref{E-imperf-test}.
The parameter estimates are in \cref{SEIR:table:ec-params}.
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\kappa_Y$ & $\kappa_{X1}$ & $\kappa_{X2}$ & $p_x$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for $P$, the EC distribution fit to the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

To link the model with prevalence measurements, I use a beta-binomial likelihood.
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests on days $t$, then the likelihood is:
\begin{align}
    \prod_{i,t} \BB (y_{it} \mid n_{it}, \sum_{k=1}^l p_{ki}(t), \rho)
\end{align}
using the mean-dispersion parametrisation of the beta-binomial distribution.
$\rho$ is a nuisance parameter controlling the dispersion of the beta-binomial distribution.

@@

\begin{figure}
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

    \node[state] (S) {$s$};
    \node[state, right=of S] (E1) {$e_1$};
    \node[state, right=of E1] (E2) {$e_2$};
    \node[state, right=of E2] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2] (R) {$r$};

    \path (S) edge node {$\sum_j \beta_{ij} i_j$} (E1)
          (E1) edge node {$\sigma$} (E2)
          (E2) edge node {$\sigma$} (I1)
          (I1) edge node {$\gamma$} (I2)
          (I2) edge node {$\gamma$} (R);

    \node[state, below=3cm of E1] (P0) {$p_0$};
    \node[state, right=of P0] (P1) {$p_1$};
    \node[state, right=of P1] (P2) {$p_2$};
    \node[state, right=of P2] (P3) {$p_3$};
    \node[state, right=of P3, draw=none] (P4) {};

    \path  (S) edge node {$\sum_j \beta_{ij} i_j$} (P0)
          (P0) edge node {$\kappa_0$} (P1)
          (P1) edge node {$\kappa_Y$} (P2)
          (P2) edge node {$p_x \kappa_{X1}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - p_x) \kappa_{X1}$} (P4)
          (P3) edge node {$\kappa_{X2}$} (P4);

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single strata $i$.}
\end{figure}
